The Multi-Stencil Language

### 2016-05-30 ARTICLES/IJPP/article/article.tex
IJPP article conclusion added
IJPP article parallelism.tex modifications

%\newtheorem{theorem}{Theorem}[section]
\newtheorem{myprop}[theorem]{Proposition}
% \newenvironment{proof}[1][Proof]{\begin{trivlist}
% \item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{mydef}[1][Definition]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{myth}[1][Theorem]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

@@

\author{Helene Coullon \and
        Julien Bigot \and
        Christian Perez
}

@@

\institute{Helene Coullon \and Christian Perez \at
              INRIA, ENS Lyon, 46 Allée d'Italie, 69007 Lyon\\
              \email{helene.coullon@inria.fr}
           \and
           Julien Bigot \at
           Maison de la Simulation, CEA, CNRS, Univ. Paris-Sud, UVSQ, Universit\'e Paris-Saclay, 91191 Gif-sur-Yvette, France\\
           \email{julien.bigot@cea.fr}
}

@@

\begin{abstract}
Insert your abstract here.
\keywords{First keyword \and Second keyword \and More}
\end{abstract}

%------------------------------------------------------------------------------
\section{Introduction}
\label{sect:introduction}
\input{introduction.tex}
%------------------------------------------------------------------------------
\section{Related work}
\label{sect:rel}
\input{related.tex}
%------------------------------------------------------------------------------
\section{Computational model of Multi-Stencil Programs}
\label{sect:formalism}
\input{formalism.tex}
%------------------------------------------------------------------------------
\section{The Multi-Stencil Language}
\label{sect:msl}
\input{msl.tex}
%------------------------------------------------------------------------------
\section{Parallelization of Multi-Stencil Programs}
\label{sect:parallelism}
\input{parallelism.tex}
%------------------------------------------------------------------------------
\section{Static Scheduling and performance model}
\label{sect:msp}
\input{msp.tex}
%------------------------------------------------------------------------------
\section{Implementation}
\label{sect:comp}
\input{compiler.tex}
%------------------------------------------------------------------------------
\section{Evaluation}
\label{sect:eval}
\input{evaluation.tex}
%------------------------------------------------------------------------------
\section{Conclusion}
\label{sect:concl}
%------------------------------------------------------------------------------

>>>>

%\newtheorem{theorem}{Theorem}[section]
\newtheorem{myprop}[theorem]{Proposition}
\newtheorem{mydef}[theorem]{Definition}
\newtheorem{myth}[theorem]{Theorem}
% \newenvironment{proof}[1][Proof]{\begin{trivlist}
% \item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
%\newenvironment{mydef}[1][Definition]{\begin{trivlist}
%\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
%\newenvironment{myth}[1][Theorem]{\begin{trivlist}
%\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{mydefs}[1][Definitions]{\par\medskip\noindent\textbf{#1}\nopagebreak\begin{itemize}[noitemsep,topsep=.25\topsep]}{\end{itemize}}

@@

\author{H\'el\`ene Coullon \and Julien Bigot \and Christian Perez}

@@

\institute{H\'el\`ene Coullon \and Christian Perez \at
              INRIA, LIP, ENS Lyon, 46 Allée d'Italie, 69364 Lyon Cedex 07, France\\
              \email{helene.coullon@inria.fr}
           \and
           Julien Bigot \at
           Maison de la Simulation, CEA, CNRS, Univ. Paris-Sud, UVSQ, Universit\'e Paris-Saclay, 91191 Gif-sur-Yvette, France\\
           \email{julien.bigot@cea.fr}
}

@@

\begin{abstract}
As the computation power of modern high performance architectures increases, their heterogeneity and complexity also become more important. One of the big challenges of exascale is to get programming models which gives access to high performance computing (HPC) to many scientists and not only to a few HPC specialists. One relevant solution to ease parallel programming for scientists is Domain Specific Language (DSL). However, one problem to avoid with DSLs is to not design a new DSL each time a new domain or a new problem has to be solved. This phenomenon happens for stencil-based numerical simulations, for which a large number of languages has been proposed without code reuse between them. The Multi-Stencil Language (MSL) presented in this paper is a language common to any kind of mesh used into a stencil-based numerical simulation. It is said that MSL is mesh-agnostic. Actually, from the description of a numerical simulation, MSL produces an empty parallel pattern, or skeleton, of the simulation which will be filled using other existing parallel languages and libraries. Thus, MSL, by finding a common language for different kinds of stencil-based simulation, facilitates code reuse. MSL is evaluated on a real case simulation which solves shallow-water equations. It is shown that MSL does not introduce overheads on data parallelism up to 16.384 cores, and that the hybrid parallelism (data and task) introduced improves performance of the simulation.

\keywords{Domain Specific Language (DSL) \and stencil \and numerical simulation \and mesh-agnostic \and data parallelism \and task parallelism \and scheduling \and MPI \and OpenMP}
\end{abstract}

%------------------------------------------------------------------------------
\section{Introduction}
\label{sect:introduction}
\input{introduction.tex}
%------------------------------------------------------------------------------
\section{Related work}
\label{sect:rel}
\input{related.tex}
%------------------------------------------------------------------------------
\section{Computational model of Multi-Stencil Programs}
\label{sect:formalism}
\input{formalism.tex}
%------------------------------------------------------------------------------
\section{The Multi-Stencil Language}
\label{sect:msl}
\input{msl.tex}
%------------------------------------------------------------------------------
\section{Parallelization of Multi-Stencil Programs}
\label{sect:parallelism}
\input{parallelism.tex}
%------------------------------------------------------------------------------
\section{Static Scheduling and performance model}
\label{sect:msp}
\input{msp.tex}
%------------------------------------------------------------------------------
\section{Implementation}
\label{sect:comp}
\input{compiler.tex}
%------------------------------------------------------------------------------
\section{Evaluation}
\label{sect:eval}
\input{evaluation.tex}
%------------------------------------------------------------------------------
\section{Conclusion}
\label{sect:concl}
\input{conclusion.tex}
%------------------------------------------------------------------------------

### 2016-05-30 ARTICLES/IJPP/article/msl.tex

\begin{filecontents*}{grammar.txt}
program ::= "mesh:" meshid 
            "mesh entities:" listgroup
            "computation domains:" listcompdom
            "independent:" listinde
            "stencil shapes:" liststencil
            "quantities:" listquantities
            "scalar" listscalar
            listloop

@@

\begin{filecontents*}{exmsl.txt}
mesh : cart
mesh entities : cell, edgex
computation domains :
  d1 in cell
  d2 in edgex
independent :
  d1 and d2
stencil shapes : 
  ncc from cell to cell
  nce from cell to edgex
  nec from edgex to cell
quantities :
  cell A,B,D,E,F,G,I,J
  edgex C,H
scalar : mu, tau
time : 500
computations :
  B[d1] = k0(tau,A)
  C[d2] = k1(B[nec])
  D[d1] = k2(C)
  E[d1] = k3(C)
  F[d1] = k4(D,C[nce])
  G[d1] = k0(mu,tau,E)
  H[d2] = k6(F)
  I[d1] = k7(G,H)
  J[d1] = k8(mu,I[ncc])
\end{filecontents*}

@@

\item The \texttt{quantities} keyword (Fig.\ref{fig:grammar},~l.6) introduces identifiers for $\delta\in\Delta$, the quantities with the group of entities they are mapped on (Fig.\ref{fig:grammar},~l.16). For example the quantities \texttt{C} and \texttt{H} are mapped on \texttt{edgex} entities.

\item The \texttt{scalar} keyword (Fig.\ref{fig:grammar},~l.7) introduces identifiers for $s\in\mathcal{S}$, the scalars. For example \texttt{mu} and \texttt{tau} in Fig.\ref{fig:exmsl},~l.15.

>>>>

\begin{filecontents*}{grammar.txt}
program ::= "mesh:" meshid 
            "mesh entities:" listgroup
            "computation domains:" listcompdom
            "independent:" listinde
            "stencil shapes:" liststencil
            "mesh quantities:" listquantities
            "scalars" listscalar
            listloop

@@

\begin{filecontents*}{exmsl.txt}
mesh : cart
mesh entities : cell, edgex
computation domains :
  d1 in cell
  d2 in edgex
independent :
  d1 and d2
stencil shapes : 
  ncc from cell to cell
  nce from cell to edgex
  nec from edgex to cell
mesh quantities :
  cell A,B,D,E,F,G,I,J
  edgex C,H
scalars : mu, tau
time : 500
computations :
  B[d1] = k0(tau,A)
  C[d2] = k1(B[nec])
  D[d1] = k2(C)
  E[d1] = k3(C)
  F[d1] = k4(D,C[nce])
  G[d1] = k0(mu,tau,E)
  H[d2] = k6(F)
  I[d1] = k7(G,H)
  J[d1] = k8(mu,I[ncc])
\end{filecontents*}

@@

\item The \texttt{mesh quantities} keyword (Fig.\ref{fig:grammar},~l.6) introduces identifiers for $\delta\in\Delta$, the quantities with the group of entities they are mapped on (Fig.\ref{fig:grammar},~l.16). For example the quantities \texttt{C} and \texttt{H} are mapped on \texttt{edgex} entities.

\item The \texttt{scalars} keyword (Fig.\ref{fig:grammar},~l.7) introduces identifiers for $s\in\mathcal{S}$, the scalars. For example \texttt{mu} and \texttt{tau} in Fig.\ref{fig:exmsl},~l.15.

### 2016-05-30 ARTICLES/IJPP/article/related.tex

Many domain specific languages have been proposed for stencil computations. Each one has its own specificities and answers to a specific stencil case or to a specific additionnal optimization. For example, applied on Cartesian meshes, Pochoir~\cite{spaaTangCKLL11} works on cache optimization techniques for stencils applied onto Cartesian meshes. On the other hand, PATUS~\cite{citeulike12258902} proposes to add a parallelization strategy grammar to its stencil language to perform an auto-tuning parallelization strategy. Moreover, Halide~\cite{Ragan-Kelley:2013:HLC:2491956.2462176}  proposes an optimization and parallelization of a pipeline of stencil codes, while ExaSlang~\cite{Schmitt:2014:EDL:2691166.2691171} is specific to multigrid numerical methods etc.

The reason why each of those languages is implemented from scratch to answer its own particularities and its own optimizations is because each language is built and thought as a single block, which makes impossible or difficult code reuse from one language to another. In other words domain specific languages (for stencils or other domains) suffer from a lack of software engineering properties, which could increase the productivity to build a new language, by code-reuse, and also the maintainability of languages, with more separation of concerns. A few works propose frameworks to create DSL, which afterwards will be easier to compose and combine together~\cite{Sujeeth:2013:CRC:2524984.2524988}. The Mulsi-Stencil Language (MSL), presented in this paper, is itself a domain specific language for stencil-based simulations. It offers a way to give a mesh-agnostic description of a simulation, which could be common to different cases of simulations, and thus facilitates reuse of existing underlying data structures, optimizations or parallelization techniques. In other words, MSL extracts where parallelization and optimizations are needed into the simulation, by producing an empty parallel pattern, but not how to do it, which is left to existing languages (SkelGIS and OpenMP in this paper), and which is an implementation concern.

@@

First, the formalism is more flexible in MSL. For example, a computation can be applied onto a subset of the space domain which is not possible with Liszt and Nabla. This functionnality is an important flexibility in numerical simulations yet. For example, many computations performed in the simulation studied in this paper have to be applied onto a subpart of the overall space domain.

Second, in Liszt and Nabla the description of a simulation is not splitted from its implementation concerns. In other words, the topology of a mesh as well as numerical codes are given at the same time than the description of the different computations to apply into the simulation, while MSL splits those two different phases. Thus, MSL improves separation of concerns and maintainability of a given simulation. Actually, a numerician could describes the different computations to perform into the simulation, while another one, later, could focuss on the implementation and the numerical code of the simulation. This also facilitates reuse of other languages.

>>>>

Many domain specific languages have been proposed for stencil computations. Each one has its own specificities and answers to a specific stencil case or to a specific additionnal optimization. For example, Pochoir~\cite{spaaTangCKLL11} works on cache optimization techniques for stencils applied onto Cartesian meshes. On the other hand, PATUS~\cite{citeulike12258902} proposes to add a parallelization strategy grammar to its stencil language to perform an auto-tuning parallelization strategy. Moreover, Halide~\cite{Ragan-Kelley:2013:HLC:2491956.2462176}  proposes an optimization and parallelization of a pipeline of stencil codes, while ExaSlang~\cite{Schmitt:2014:EDL:2691166.2691171} is specific to multigrid numerical methods etc.

The reason why each of those languages are implemented from scratch to answer its own particularities and its own optimizations is because each language is built and thought as a single block, which makes impossible or difficult code reuse from one language to another. In other words domain specific languages (for stencils or other domains) suffer from a lack of software engineering properties, which could increase the productivity to build a new language, by code-reuse, and also the maintainability of languages, with more separation of concerns. A few works propose frameworks to create DSL, which afterwards will be easier to compose and combine together~\cite{Sujeeth:2013:CRC:2524984.2524988}. The Mulsi-Stencil Language (MSL), presented in this paper, is itself a domain specific language for stencil-based simulations. It offers a way to give a mesh-agnostic description of a simulation, which could be common to different cases of simulations, and thus facilitates reuse of existing underlying data structures, optimizations or parallelization techniques. In other words, MSL extracts where parallelization and optimizations are needed into the simulation, by producing an empty parallel pattern, but not how to do it, which is left to existing languages (SkelGIS and OpenMP in this paper), and which is an implementation concern.

@@

First, the formalism is more flexible in MSL. For example, a computation can be applied onto a subset of the space domain which is not possible with Liszt and Nabla. This functionnality is important in numerical simulations yet. For example, many computations performed in the simulation studied in this paper have to be applied onto a subpart of the overall space domain.

Second, in Liszt and Nabla the description of a simulation is not splitted from its implementation concerns. In other words, the topology of a mesh as well as numerical codes are given at the same time than the description of the different computations to apply into the simulation, while MSL splits those two different phases. Thus, MSL improves separation of concerns and maintainability of a given simulation. Actually, a numerician could describes the different computations to perform into the simulation, while another one, later, could focuss on the implementation and the numerical code. This also facilitates reuse of other languages.
